The sweet spot for local polygenic adaptation: a case study motivated by stickleback

### 2018-07-18 Stickleback_Paper.tex
Update on Overleaf.

\title{The sweet spot for local polygenic adaptation: a case study motivated by stickleback}
\author{Jared Galloway, Bill Cresko, and Peter Ralph}
\maketitle

@@

@@

Ninespine stickleback provide one of the most striking examples of local adaptation,
as the hemisphere-wide metapopulation includes both marine populations
and a large number of smaller freshwater populations
that have apparently repeatedly adapted to freshwater conditions
often using the same genetic basis.
In this paper, we use simulations motivated by stickleback populations
to examine what amounts of gene flow favor stable metapopulation polymorphism
with allele sharing,
and to further dissect the underlying dynamics.
We find that rapid, repeated adaptation 
using alleles maintained at low frequency by migration-selection balance
(the ``transporter hypothesis'')
occurs over a realistic range of intermediate rates of gene flow,
between slow, independent adaptation with low gene flow
and large migration load at high gene flow.
This is mediated mainly by the total downstream influx of alleles \emph{(maybe??)}.
We do not see evidence for strong effects encouraging genomic clustering of causal alleles
or towards particular dominance coefficients.
$F_{ST}$ scans for adaptive alleles are more likely to succeed with higher rates of gene flow,
The results support existing theory of local adaptation,
and provide a more concrete look at a particular, empirically motivated example.

@@

Recently, multiple instances of similar (parallel) underlying genetic basis of rapid local adaptation has brought about 
questions surrounding the origins and maintenance of genetic polymorphism.
A empirical model is the Alaskan populations of freshwater and marine Ninespine Stickleback fish.
In 1964, The Great Alaskan Earthquake caused an uplift of Middleton island and in turn,
introduced a group of freshwater ponds around the perimeter of the island. 
Quickly inhabited by the surrounding marine population of Stickleback,
\citet{lescak2015evolution} observed significant phenotypic changes in less than 50 years that appear to be 
parallel to freshwater stickleback that have been separated for over $13,000$ years.
In freshwater stickleback, the number of lateral plates are reduced
and the opercle shapes shows the same expansion of the dorsal region and reduction of the ventral region.
\jgg{Bill - more specifics about Bill, Suzie, Kristin  \& or Thom's papers .. ? }
\bill{no way}
These results leave us with questions surrounding the nature of rapid adaptation.
Does convergent evolution breed it's own solution (haplotype) for every new selective pressure, 
or can these solutions can be efficiently shared across multiple sub - populations facing similar selective pressures.

@@

We explored these questions
using forwards-time simulations with explicit genomic representation of a quantitative trait
in SLiM \citep{haller2017slim}.
The details (described below) are motivated by the nine-spine stickleback,
but remain simplistic in some aspects due to computational constraints.
Possibly the most important caveat is that simulated population sizes are much smaller
than the census size of the nine-spine stickleback population;
we will consider the likely effects of this in the Discussion.
\plr{TODO}

\paragraph{Habitat and geography}
Our model has two habitat types, which are separate populations
with a fixed total size of 2,000 diploid individuals each.
The scheme, depicted in Figure \ref{fig:Geo},
roughly models a set of freshwater habitats
along a stretch of coastline.
The ``marine'' habitat  is a continuous, one-dimensional range of 10 units of length,
and the ``freshwater'' habitat is divided into ten subpopulations
(which we call ``lakes''), each connected to the marine habitat
at positions $i - 1/2$ for $1 \le i \le 10$.

Fitness in each habitat is determined by a single continuous trait with an inherent tradeoff,
to roughly model the cumulative effect of the various phenotypes
such as lateral plating and opercle shape
on which divergent selection acts in the two environments.
The optimal trait value in the marine habitat is $+10$,
and in the freshwater habitat is $-10$,
and fitness of a fish with trait value $x$ in a habitat with optimal value $x_\text{opt}$ is
Gaussian with standard deviation, i.e.,
\begin{align*}
    f(x; x_0)
    &=
    \exp\left\{
        \frac{1}{2}
            \left(
            \frac{x-x_\text{opt}}{15}
            \right)^2
        \right\} .
\end{align*}

@@

\paragraph{Potential genomic architecture}
Each individual carries two copies of a linear chromosome of $10^5$ loci
(which could be thought of roughly as a kilobase each).
New mutations occur with probability $10^{-7}$ per locus per generation.
Ten ``effect regions'' of 100 loci each
are spread evenly along the chromosome (separated by 4950 loci)
in which mutations affect the trait;
mutations in other parts of the genome are neutral.
Each mutation in these regions is either additive, completely recessive,
or completely dominant (with equal probability),
and with an effect sizes chosen randomly from an Exponential distribution with mean $1/2$,
either positive or negative with equal probability.
Individual trait value are determined additively from the diploid genotypes.
Concretely, an individual that is heterozygous and homozygous for mutations
at loci $H$ and $D$ respectively
has trait value
$x = \sum_{\ell \in H} h_\ell s_\ell + \sum_{\ell \in D} s_\ell$,
where $s_\ell$ is the effect size of the mutation at locus $\ell$.
Subsequent mutations at the same locus replace the previous allele.

@@

Migration occurs both locally along the coastline in the marine habitat
and between the marine habitat and the lakes,
and can be thought of as occurring at the juvenile stage.
The between-habitat migration rate is denoted $m$,
and will be called simply the ``migration rate''.
Each new individual in the marine habitat has freshwater parents with probability $m$;
to obtain the pair, a first parent is chosen proportional to fitness,
and a mate is chosen from the same lake as the first, also proportional to fitness.
The resulting offspring is given a spatial location in the marine habitat
at the location of the parent's lake.
Parents for a new marine individual who is not a migrant are chosen similarly (with probability $1-m$):
first, a single parent is chosen proportionally to fitness in the marine habitat,
and then a mate is chosen, 
also proportionally to fitness but reweighted by a Gaussian function 
of the distance separating the two, with stanadard deviation $1/2$: 
concretely, if the first parent is marine individual $i$,
then marine individual $j$ is chosen as the mate
with probability proportional to $f(x_i) \exp(-2d_{ij}^2)$,
where $d_{ij}$ is the distance between the two locations.
Finally, each new marine offspring is given a position 
displaced from the first parent's position by a random Gaussian distance
with mean 0 and standard deviation 0.02, and reflected to stay within the population.
New offspring in the freshwater habitat are chosen in the same way,
except the probability that the parents are marine individuals is $m$;
any new freshwater offpsring produced by marine individuals
are assigned to the lake nearest to the position of the first marine parent.

\paragraph{New lakes}
To study adaptation in newly appearing habitats,
we introduce a new set of lakes midway through the simulation.
The initial set of individuals have parents chosen as above from the marine habitat,
and act like an independent copy of the original set of lakes -- in particular,
the two sets of lakes each have 2,000 individuals at all times.
Since this doubles the number of lake-to-marine immigrants,
after this happens the probability that a new marine individual has freshwater parents
is $2m$ instead of $m$.

\paragraph{Descriptive statistics}
To assess whether new lakes adapt using existing genetic diversity,
we define a \emph{pre-existing freshwater adapted allele}
to be an effect mutation that has frequency higher than $0.5$ in at least one of the original lakes,
while remaining lower than $0.5$ in the marine. 
This categorization is made for each generation using the allele frequencies from that generation,
and so changes with time.
Alleles common in the newly introduced lakes do not count
if they are not also common in the original lakes.
They are defined this way because the transportation hypothesis 
does not specify where or when an advantageous mutation arises,
but simply suggests that any sufficiently common freshwater adapted allele 
could participate in adaptaiton in new habitat \citep{schluter2009genetics}.

>>>>

\title{The sweet spot for local polygenic adaptation: a case study motivated by stickleback}
\author{Jared Galloway, William A. Cresko, and Peter Ralph}
\maketitle

@@

\bill{A simulation study of gene flow and local polygenic adaptation in stickleback fish}

@@

Threespine stickleback fish provide one of the most striking examples of local adaptation. This hemisphere-wide metapopulation includes both marine populations and a large number of smaller freshwater populations
that have apparently repeatedly adapted to freshwater conditions
often using the same genetic basis.
In this paper, we use simulations motivated by stickleback populations
to examine what amounts of gene flow favor stable metapopulation polymorphism
with allele sharing,
and to further dissect the underlying dynamics.
We find that rapid, repeated adaptation 
using alleles maintained at low frequency by migration-selection balance
(the ``transporter hypothesis'')
occurs over a realistic range of intermediate rates of gene flow,
between slow, independent adaptation with low gene flow
and large migration load at high gene flow.
This is mediated mainly by the total downstream influx of alleles \emph{(maybe??)}.
We do not see evidence for strong effects encouraging genomic clustering of causal alleles
or towards particular dominance coefficients.
$F_{ST}$ scans for adaptive alleles are more likely to succeed with higher rates of gene flow,
The results support existing theory of local adaptation,
and provide a more concrete look at a particular, empirically motivated example.

@@

Recently, multiple instances of similar (parallel) underlying genetic basis of rapid local adaptation has brought about 
questions surrounding the origins and maintenance of genetic polymorphism.
A empirical model is the Alaskan populations of freshwater and marine Ninespine Stickleback fish.
In 1964, The Great Alaskan Earthquake caused an uplift of Middleton island and in turn,
introduced a group of freshwater ponds around the perimeter of the island. 
Quickly inhabited by the surrounding marine population of Stickleback,
\citet{lescak2015evolution} observed significant phenotypic changes in less than 50 years that appear to be 
parallel to freshwater stickleback that have been separated for over $13,000$ years.
In freshwater stickleback, the number of lateral plates are reduced
and the opercle shapes shows the same expansion of the dorsal region and reduction of the ventral region.
% \jgg{Bill - more specifics about Bill, Suzie, Kristin  \& or Thom's papers .. ? }
These results leave us with questions surrounding the nature of rapid adaptation.
Does convergent evolution breed it's own solution (haplotype) for every new selective pressure, 
or can these solutions can be efficiently shared across multiple sub - populations facing similar selective pressures.

@@

We explored these questions
using forwards-time simulations with explicit genomic representation of a quantitative trait
in SLiM \citep{haller2017slim}.
The details (described below) are motivated by previous population genomic studies of threespine stickleback,
but remain simplistic in some aspects due to computational constraints.
Possibly the most important caveat is that simulated population sizes are much smaller
than the census size of the threespine stickleback population;
we will consider the likely effects of this in the Discussion.
\plr{TODO}

\paragraph{Habitat and geography}
Our model has two habitat types, which are considered separate populations
with a fixed total size of 2,000 diploid individuals each. The marine population is treated as a single large deme, but the freshwater population is subdivided into local subpopulations or demes.
The scheme, depicted in Figure \ref{fig:Geo},
roughly models a set of freshwater habitats
along a stretch of coastline.
The ``marine'' habitat  is a continuous, one-dimensional range of 10 units of length,
and the ``freshwater'' habitat is divided into ten subpopulations
(which we call ``lakes''), each connected to the marine habitat
at positions $i - 1/2$ for $1 \le i \le 10$.

Fitness in each habitat is determined by a single continuous trait with an inherent tradeoff such that alternative values of the trait have high or low fitness in tht two different habitats. This situation roughly models the cumulative effect of the various phenotypes
such as armor morphology, body size, craniofacial variation and opercle shape (among others)
on which divergent selection acts in the two environments.
The optimal trait value in the marine habitat is $+10$,
and in the freshwater habitat is $-10$. Fitness of a fish with trait value $x$ in a habitat with optimal value $x_\text{opt}$ is
Gaussian with standard deviation, i.e.,
\begin{align*}
    f(x; x_0)
    &=
    \exp\left\{
        \frac{1}{2}
            \left(
            \frac{x-x_\text{opt}}{15}
            \right)^2
        \right\} .
\end{align*}

@@

\paragraph{Potential genomic architecture}\bill{Initial genomic architecture? The header is a little confusing to me}
Each individual carries two copies of a linear chromosome of $10^5$ loci
(which could be thought of roughly as a kilobase each).
New mutations occur with probability $10^{-7}$ per locus per generation.
Ten ``effect regions'' of 100 loci each
are spread evenly along the chromosome (separated by 4950 loci)
in which mutations affect the trait;
mutations in other parts of the genome are neutral.
Each mutation in these regions is either additive, completely recessive,
or completely dominant (with equal probability),
and with an effect sizes chosen randomly from an Exponential distribution with mean $1/2$,
either positive or negative with equal probability.
Individual trait value are determined additively from the diploid genotypes.
Concretely, an individual that is heterozygous and homozygous for mutations
at loci $H$ and $D$ respectively
has trait value
$x = \sum_{\ell \in H} h_\ell s_\ell + \sum_{\ell \in D} s_\ell$,
where $s_\ell$ is the effect size of the mutation at locus $\ell$.
Subsequent mutations at the same locus replace the previous allele.

@@

Migration occurs both locally along the coastline in the marine habitat
and between the marine habitat and the lakes,
and can be thought of as occurring at the juvenile stage.
The between-habitat migration rate is denoted $m$,
and will be called simply the ``migration rate''.
Each new individual in the marine habitat has freshwater parents with probability $m$;
to obtain the pair, a first parent is chosen proportional to fitness,
and a mate is chosen from the same lake as the first, also proportional to fitness.
The resulting offspring is given a spatial location in the marine habitat
at the location of the parent's lake.
Parents for a new marine individual who is not a migrant are chosen similarly (with probability $1-m$):
first, a single parent is chosen proportionally to fitness in the marine habitat,
and then a mate is chosen, 
also proportionally to fitness but reweighted by a Gaussian function 
of the distance separating the two, with stanadard deviation $1/2$. 
Concretely, if the first parent is marine individual $i$,
then marine individual $j$ is chosen as the mate
with probability proportional to $f(x_i) \exp(-2d_{ij}^2)$,
where $d_{ij}$ is the distance between the two locations.
Finally, each new marine offspring is given a position 
displaced from the first parent's position by a random Gaussian distance
with mean 0 and standard deviation 0.02, and reflected to stay within the population.
New offspring in the freshwater habitat are chosen in the same way,
except the probability that the parents are marine individuals is $m$;
any new freshwater offpsring produced by marine individuals
are assigned to the lake nearest to the position of the first marine parent.

\paragraph{New lakes}
To study adaptation in newly appearing freshwater habitats,
we introduce a new set of lakes midway through the simulation.
The initial set of individuals have parents chosen as above from the marine habitat,
and act like an independent copy of the original set of lakes -- in particular,
the two sets of lakes each have 2,000 individuals at all times.
Since this doubles the number of lake-to-marine immigrants,
after this happens the probability that a new marine individual has freshwater parents
is $2m$ instead of $m$.

\paragraph{Descriptive statistics}\bill{Population genetic analyses? Instead of the more general 'Descriptive statistics'?}
To assess whether new lakes adapt using existing genetic diversity,
we define a \emph{pre-existing freshwater adapted allele}
to be an effect mutation that has frequency higher than $0.5$ in at least one of the original lakes,
while remaining lower than $0.5$ in the marine. 
This categorization is made for each generation using the allele frequencies from that generation,
and so changes with time.
Alleles common in the newly introduced lakes do not count
if they are not also common in the original lakes.
They are defined this way because the transportation hypothesis 
does not specify where or when an advantageous mutation arises,
but simply suggests that any sufficiently common freshwater adapted allele 
could participate in adaptaiton in new habitat \citep{schluter2009genetics}.
